---
title: "DCS: Phase Space Reconstruction and RQA"
author: "[Fred Hasselman](mailto:f.hasselman@pwo.ru.nl) & [Maarten Wijnants](mailto:m.wijnants@pwo.ru.nl)"
date: '2015-2016'
output:
  html_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 5
    highlight: pygments
    theme: spacelab
    toc: yes
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 5
    fig_width: 7
    highlight: zenburn
    keep_tex: yes
    latex_engine: xelatex
    toc: yes
    toc_depth: 4
monofont: Gill Sans Light
mainfont: Calibri
sansfont: Gill Sans Light
geometry: a4paper
---
```{r set-options, echo=FALSE,include=FALSE}
require(knitr)
require(formatR)
require(dplyr)
options(width=200)
knitr::opts_chunk$set(cache=FALSE,prompt=FALSE,comment=">",message=FALSE,echo=TRUE,warning=FALSE,tidy=TRUE,strip.white=TRUE,size="small", fig.align = "center",fig.show='hold')

```

# Phase Space Reconstruction: Quantifying Recurrent Patterns

You can use `R` or `Matlab` to run `RQA` analyses. This document is based on `R`. Yoiu can find the assignments for `Matlab` on Blackboard.

### Using R: Install package `nlRtsa`

First, download an install package `nlRtsa` (under construction)

> IMPORTANT!: 
> Make sure you install the package to a library in directory in which you have write access. 
> In the computer rooms at the university, the easiest way to make sure things run smoothly is to first use the `RStudio` menu on the `Packages` pane to install a package, e.g., `devtools`. `RStudio` will give you the option to install the package in your personal user directory.

```{r,eval=FALSE}
require(devtools)
install_github("FredHasselman/nlRtsa")
library(nlRtsa)
```

If this does not work, run the following code:
```{r,evaluate = FALSE}
require(devtools)
source_url("https://raw.githubusercontent.com/FredHasselman/nlRtsa/master/filesource/nlRtsa_SOURCE.R")
```

This will source all the functions in `nlRtsa` into your environment.    
Drawback is that you cannot access the manual pages.

## Assignment: Phase Space Reconstruction

We'll package `fractal` to reconstruct some phase spaces.

* Install and load these packages by running: `in.IT('fractal','rgl')` 
    + Run the folowing code: `plot3d(lorenz)`
* Package fractal includes the 3 dimensions of the Lorenz system in the chaotic regime. We'll reconstruct it based on just dimension `X`
* Use `lx <- lorenz[1:2048,1]` to reconstruct the phase space. 
        + Find an optimal embedding lag using `timeLag`, use `method = "mutual"`.
        + Find the embeddingdimension, using `FNN`
        + Plot the phase space (use `open3d()` to open an interactive plot if you like.
