---
title: "DCS: "
author: "[Fred Hasselman](mailto:f.hasselman@pwo.ru.nl) & [Maarten Wijnants](mailto:m.wijnants@pwo.ru.nl)"
date: '2015-2016'
output:
  html_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 5
    highlight: pygments
    theme: spacelab
    toc: yes
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 5
    fig_width: 7
    highlight: zenburn
    keep_tex: yes
    latex_engine: xelatex
    toc: yes
    toc_depth: 4
monofont: Gill Sans Light
mainfont: Calibri
sansfont: Gill Sans Light
geometry: a4paper
---
```{r set-options, echo=FALSE,include=FALSE}
require(knitr)
require(formatR)
require(dplyr)
options(width=200)
knitr::opts_chunk$set(cache=FALSE,prompt=FALSE,comment=">",message=FALSE,echo=TRUE,warning=FALSE,tidy=TRUE,strip.white=TRUE,size="small", fig.align = "center",fig.show='hold')

```

# Fractal Geometry: Estimating Self-Affinity and Dimension of Time Series

It will become increasingly difficult to use software like Excel and SPSS. Perhaps now is a good time to switch to R or Matlab. We do have a spreadsheet example of Standardised Dispersion Anaysis. 

### Using R: Install package `nlRtsa`

First, download an install package `nlRtsa` (under construction)

```{r}
require(devtools)
install_github("FredHasselman/nlRtsa")
library(nlRtsa)
```

### Examples: Fast Fourier transform and Power Spectrum

The `nlRtsa` package contains a nice function `in.IT()` that will load a list of packages and install them if they are not present on your system.
```{r}
in.IT(c("signal","pracma"))
```

Below is an example of a signal built from sine components (`y`) whose relative amplitudes are recovered in the powerspectrum.
```{r}
# Sawtooth
x <- seq(-3.2,3.2, length.out = 256)
y <- 2*sin(10*x) - 1*sin(20*x) + (2/3)*sin(30*x) - (1/2)*sin(40*x) + (2/5)*sin(50*x) - (1/4)*sin(60*x)

# Plot the sawtooth wave as constructed by the Fourier series above
plot(x,y, xlab ='Time (a.u.)', ylab = 'Variable (a.u.)', main ='Sawtooth wave', type = "l")

# Perform a Fast Fourier Transform and calculate the Power and Frequency
Y <- fft(y)
Pyy <- Y*Conj(Y)/256
f <- 1000/256*(0:127)

# Plot the power spectrum of the sawtooth wave
plot(f[1:50],Pyy[1:50], type="b",xlab='Frequency (a.u.)', ylab ='Power (a.u.)', pch=21, bg='grey60', main = 'Power Spectrum')
```

Now we do the same for a very noisy signal into which we insert a frequency.
```{r}
# A time vector
t <- pracma::linspace(x1 = 0, x2 = 50, n = 256)
# There are three sine components
x <- sin(2*pi*t/.1) + sin(2*pi*t/.3) + sin(2*pi*t/.5)
# Add random noise!
y <- x + 1*randn(size(t))

# Plote the noise.
plot(t, y, type = "l", xlab = 'Time (a.u.)', ylab = 'Variable (a.u.)', main = 'A very noisy signal')

# Get the frequency domain
Y <- fft(y)
Pyy <- Y*Conj(Y)/256
f <- 1000/256*(0:127)

# Plot the power spectrum of this noise
plot(f[1:50],Pyy[1:50], type="b",xlab='Frequency (a.u.)', ylab='Power (a.u.)', pch=21, bg='grey60', main = 'Power Spectrum')
```

## Assignment: The Spectral Slope

We can use the power spectrum to estimate a **self-affinity parameter**, or scaling exponent.

* Download `ts1.txt`, `ts2.txt`, `ts3.txt` from blackboard. The text files contain time series.
* Load the three time series into `R` using any of the `read` functions. **TIP** Use the function `import()` in package `rio`.
* Plot the three time series.

Now we can do some basic data preparations:

* Are the lengths of the time series a power of 2? (Use `log2(length of var)` )
* Are the data normalized? (we will not remove datapoints outside 3SD)
    + To normalize we have to subtract the mean from each value in the time series and divide it by the standard deviation, the function `scale()` can do this for you, but you could also use `mean()` and `sd()`.
* Plot the normalized time series.

Before a spectral analysis you should remove any linear trends (it cannot deal with nonstationary signals!)

* Detrend the normalized data (just the linear trend). 
    + This can be done by the command `pracma::detrend()`. 
    + Also try to figure out how to detrend the data using `stats::lm()` or `stats::poly()`
* Plot the detrended data.

The function `nlRtsa::fd.psd()` will perform the spectral slope fitting procedure. 

* Look at the manual pages to figure out how to call it.
    + Remember, we have already normalized and detrended the data.
    + You can also look at the code itself by selecting the name in`R` and pressing `F2` 
* Calculate the spectral slopes for the three normalized and detrended time series.
    + Call with `plot = TRUE`
    + Compare the results... What is your conclusion?

## Assignment: DFA and SDA

* Use the functions `fd.dfa()` and `fd.sda()` to estimate the self-affinity parameter and Dimension of the series. 
    + Think about which data preparation steps are necessary.
    + Compare the results between different methods.

## Assignment: ACF/PACF, Relative Roughness and Sample Entropy

* Also calculate the ACF, PACF, Relative Roughness (see assignment of last week) and Sample Entropy (`pracma::sample_entropy()`)
    + Compare the results.

## Assignment: Deterministic Chaos

* Generate a chaotic timeseries (e.g. $r = 4$ ) of equal length as the time series used above (use the function `growth.ac( ..., type = "logistic")` in `nlRtsa`)
* Get all the quantities for this series and compare them to the previous results.
